---
title: "Diabetes Risk Dashboard"
author: "Samridhi Purohit"
date: ""
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: scroll
    self_contained: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(tidyverse)
library(plotly)
library(scales)
library(stringr)

theme_set(theme_minimal(base_size = 14))
```

```{css, echo=FALSE}
/* Overall page */
body {
  background-color: #f6f8fb;
  color: #1f2937;
  font-family: "Helvetica Neue", Arial, sans-serif;
}

/* Top navigation bar */
.navbar,
.navbar-inverse {
  background-color: #263d52 !important;
  border-color: #263d52 !important;
}

.navbar-inverse .navbar-brand,
.navbar-inverse .navbar-nav > li > a {
  color: #ffffff !important;
}

.navbar-inverse .navbar-nav > .active > a,
.navbar-inverse .navbar-nav > .active > a:focus,
.navbar-inverse .navbar-nav > .active > a:hover {
  background-color: #0E7490 !important;
  color: #ffffff !important;
}

/* Dashboard cards */
.section.level3 {
  background-color: #ffffff;
  border: 1px solid #e5e7eb;
  border-radius: 12px;
  box-shadow: 0 2px 10px rgba(15, 23, 42, 0.06);
  margin: 12px 8px;
  overflow: hidden;
}

/* Card headers */
.section.level3 > h3 {
  background-color: #f8fafc;
  color: #0f172a;
  font-weight: 600;
  border-bottom: 1px solid #e5e7eb;
  padding: 12px 14px;
  margin: 0;
}

/* Text inside cards */
.section.level3 p {
  font-size: 16px;
  line-height: 1.5;
  padding-left: 14px;
  padding-right: 14px;
}

/* Links */
a {
  color: #0E7490;
  font-weight: 500;
}

a:hover {
  color: #174E6D;
}

/* Mobile display */
@media (max-width: 768px) {
  .navbar-brand {
    white-space: normal !important;
    line-height: 1.15;
    font-size: 22px;
  }

  .chart-scroll {
    overflow-x: auto;
    -webkit-overflow-scrolling: touch;
    width: 100%;
  }

  .chart-scroll .html-widget,
  .chart-scroll .plotly {
    min-width: 900px !important;
  }

  .modebar {
    display: none !important;
  }
}
```

```{r data_prep, include=FALSE}
# Load data
df <- read_csv("diabetes_012_health_indicators_BRFSS2015.csv", show_col_types = FALSE)

# Helper function for binary variables
yes_no <- function(x) {
  case_when(
    x == 1 ~ "Yes",
    x == 0 ~ "No",
    TRUE ~ NA_character_
  )
}

# Clean and prepare data
brfss <- df %>%
  mutate(
    diabetes_yes = if_else(Diabetes_012 == 2, 1, 0),
    
    age_group = case_when(
      Age %in% 1:2   ~ "18-29",
      Age %in% 3:5   ~ "30-44",
      Age %in% 6:7   ~ "45-54",
      Age %in% 8:9   ~ "55-64",
      Age %in% 10:13 ~ "65+",
      TRUE ~ NA_character_
    ),
    
    bmi_category = case_when(
      BMI > 0 & BMI < 18.5   ~ "Underweight",
      BMI >= 18.5 & BMI < 25 ~ "Normal",
      BMI >= 25 & BMI < 30   ~ "Overweight",
      BMI >= 30 & BMI < 35   ~ "Obese I",
      BMI >= 35              ~ "Obese II/III",
      TRUE ~ NA_character_
    ),
    
    high_bp        = yes_no(HighBP),
    high_chol      = yes_no(HighChol),
    smoker         = yes_no(Smoker),
    phys_activity  = yes_no(PhysActivity),
    heavy_alcohol  = yes_no(HvyAlcoholConsump),
    any_healthcare = yes_no(AnyHealthcare),
    no_doc_cost    = yes_no(NoDocbcCost),
    diff_walk      = yes_no(DiffWalk)
  ) %>%
  filter(!is.na(age_group), !is.na(bmi_category), BMI > 0) %>%
  mutate(
    age_group = factor(
      age_group,
      levels = c("18-29", "30-44", "45-54", "55-64", "65+")
    ),
    bmi_category = factor(
      bmi_category,
      levels = c("Underweight", "Normal", "Overweight", "Obese I", "Obese II/III")
    )
  )
```

Main Findings
=====================================

### Project links
Dashboard is available at: [https://samridhipurohit.github.io/diabetes-risk-dashboard/](https://samridhipurohit.github.io/diabetes-risk-dashboard/)  
GitHub repository: [https://github.com/samridhipurohit/diabetes-risk-dashboard](https://github.com/samridhipurohit/diabetes-risk-dashboard)

### Diabetes hotspots across age and BMI {data-height=550}

This heatmap shows where diabetes prevalence is highest across age and BMI groups. Hover over each cell to view the diabetes prevalence and sample size.

<div class="chart-scroll">

```{r heatmap_widget}
heatmap_dat <- brfss %>%
  group_by(age_group, bmi_category) %>%
  summarise(
    n = n(),
    diabetes_prev_pct = 100 * mean(diabetes_yes, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    tooltip_text = paste0(
      "Age group: ", age_group,
      "<br>BMI category: ", bmi_category,
      "<br>Diabetes prevalence: ", round(diabetes_prev_pct, 1), "%",
      "<br>Sample size: ", comma(n)
    )
  )

p1 <- ggplot(
  heatmap_dat,
  aes(
    x = bmi_category,
    y = age_group,
    fill = diabetes_prev_pct,
    text = tooltip_text
  )
) +
  geom_tile(color = "white", linewidth = 0.6) +
  labs(
    title = "Diabetes Prevalence by Age and BMI",
    x = "BMI category",
    y = "Age group",
    fill = "Diabetes prevalence (%)"
  ) +
  scale_fill_viridis_c(
  option = "cividis",
  begin = 0.15,
  end = 0.95,
  labels = function(x) paste0(round(x, 1), "%")
) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    axis.text.x = element_text(angle = 20, hjust = 1),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11)
  )

ggplotly(p1, tooltip = "text", height = 500, width = 900) %>%
  layout(
    dragmode = FALSE,
    hoverlabel = list(
      align = "left",
      bgcolor = "#f7f7f7",
      bordercolor = "#cfcfcf",
      font = list(color = "#333333")
    ),
    margin = list(l = 80, r = 120, b = 100, t = 80),
    xaxis = list(fixedrange = TRUE),
    yaxis = list(fixedrange = TRUE)
  ) %>%
  config(
    displayModeBar = FALSE,
    scrollZoom = FALSE,
    doubleClick = FALSE
  )
```
</div>

### Key takeaway

Diabetes prevalence increases consistently across both age groups and BMI categories. The highest prevalence is observed among older adults in the highest obesity category, showing where diabetes burden is most concentrated.

### Why this matters

This dashboard helps identify population subgroups where diabetes prevalence is most concentrated. These findings can support public health communication, screening priorities, and prevention strategies focused on high-risk groups.


Risk Factors
=====================================

### Risk-factor diabetes gaps among adults aged 55-64 with Obese I BMI {data-height=600}

This chart compares diabetes prevalence between people with and without each risk factor in one high-risk subgroup. Hover over each point to view the prevalence gap, factor prevalence, and sample sizes.

<div class="chart-scroll">

```{r risk_factor_widget}
selected_age <- "55-64"
selected_bmi <- "Obese I"

subgroup <- brfss %>%
  filter(age_group == selected_age, bmi_category == selected_bmi)

risk_map <- c(
  high_bp = "High blood pressure",
  high_chol = "High cholesterol",
  smoker = "Ever smoked 100 cigarettes",
  phys_activity = "Any physical activity",
  heavy_alcohol = "Heavy alcohol consumption",
  any_healthcare = "Has any healthcare coverage",
  no_doc_cost = "Could not see doctor due to cost",
  diff_walk = "Difficulty walking or climbing stairs"
)

risk_gap_dat <- purrr::map_dfr(names(risk_map), function(v) {
  
  tmp <- subgroup %>%
    transmute(
      exposure = .data[[v]],
      diabetes_yes = diabetes_yes
    ) %>%
    filter(!is.na(exposure)) %>%
    group_by(exposure) %>%
    summarise(
      n = n(),
      prev_pct = 100 * mean(diabetes_yes, na.rm = TRUE),
      .groups = "drop"
    )
  
  yes_prev <- tmp$prev_pct[tmp$exposure == "Yes"]
  no_prev  <- tmp$prev_pct[tmp$exposure == "No"]
  yes_n    <- tmp$n[tmp$exposure == "Yes"]
  no_n     <- tmp$n[tmp$exposure == "No"]
  
  tibble(
    risk_factor = risk_map[[v]],
    prev_yes_pct = ifelse(length(yes_prev) == 0, NA, yes_prev),
    prev_no_pct  = ifelse(length(no_prev) == 0, NA, no_prev),
    n_yes = ifelse(length(yes_n) == 0, NA, yes_n),
    n_no  = ifelse(length(no_n) == 0, NA, no_n),
    gap_pp = prev_yes_pct - prev_no_pct,
    factor_prev_pct = 100 * mean(subgroup[[v]] == "Yes", na.rm = TRUE)
  )
}) %>%
  mutate(
    tooltip_text = paste0(
      "Risk factor: ", risk_factor,
      "<br>Diabetes prevalence when factor = Yes: ", round(prev_yes_pct, 1), "%",
      "<br>Diabetes prevalence when factor = No: ", round(prev_no_pct, 1), "%",
      "<br>Gap: ", round(gap_pp, 1), " percentage points",
      "<br>Factor prevalence in subgroup: ", round(factor_prev_pct, 1), "%",
      "<br>N with factor: ", comma(n_yes),
      "<br>N without factor: ", comma(n_no)
    )
  ) %>%
  arrange(desc(gap_pp)) %>%
  mutate(
    highlight = ifelse(row_number() == 1, "Largest gap", "Other"),
    risk_factor = factor(risk_factor, levels = rev(unique(risk_factor)))
  )

p2 <- ggplot(
  risk_gap_dat,
  aes(
    x = gap_pp,
    y = risk_factor,
    text = tooltip_text,
    color = highlight
  )
) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#7a7a7a") +
  geom_segment(
    aes(
      x = 0,
      xend = gap_pp,
      y = risk_factor,
      yend = risk_factor
    ),
    linewidth = 1
  ) +
  geom_point(size = 3.5) +
  scale_color_manual(
  values = c("Largest gap" = "#2c7fb8", "Other" = "#CBD5E1")
) +
  labs(
    title = "Diabetes Prevalence Gaps by Risk Factor",
    x = "Gap in diabetes prevalence, percentage points",
    y = "Risk factor"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0, color = "#0f172a"),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
  )

ggplotly(p2, tooltip = "text", height = 550, width = 950) %>%
  layout(
    dragmode = FALSE,
    hoverlabel = list(
      align = "left",
      bgcolor = "#f7f7f7",
      bordercolor = "#cfcfcf",
      font = list(color = "#333333")
    ),
    margin = list(l = 260, r = 120, b = 100, t = 90),
    xaxis = list(fixedrange = TRUE),
    yaxis = list(fixedrange = TRUE)
  ) %>%
  config(
    displayModeBar = FALSE,
    scrollZoom = FALSE,
    doubleClick = FALSE
  )
```

</div>

### Key takeaway

Within adults aged 55-64 in the Obese I BMI category, high blood pressure, difficulty walking, and high cholesterol show some of the largest positive diabetes prevalence gaps. These patterns suggest that diabetes burden clusters with cardiovascular and mobility-related risk factors.

### How to interpret the chart

A positive gap means diabetes prevalence is higher among people with the risk factor compared with people without the risk factor. A negative gap means diabetes prevalence is lower among people with the risk factor compared with people without the risk factor.


Dataset
=====================================

### Dataset description

This dashboard uses the [CDC Diabetes Health Indicators dataset](https://archive.ics.uci.edu/dataset/891/cdc+diabetes+health+indicators), derived from the 2015 Behavioral Risk Factor Surveillance System. The dataset includes 253,680 survey responses from U.S. adults and contains health, lifestyle, healthcare access, and demographic indicators related to diabetes status. Data were collected through the CDC’s BRFSS, a population-based health survey. The study population includes adult survey respondents in the United States in 2015. The dashboard focuses on diabetes prevalence patterns across age, BMI, and selected health risk factors.

### Real-world impact

This dashboard helps identify population subgroups where diabetes prevalence is most concentrated. These findings can support public health communication, screening priorities, and prevention strategies focused on high-risk groups.

### GitHub repository

[View the GitHub Repository](https://github.com/samridhipurohit/diabetes-risk-dashboard)

